\documentclass[12pt]{report}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc} 
\usepackage[T1]{fontenc} 
\usepackage{lmodern} 
\usepackage{hyperref} 
\usepackage{lscape} 
\usepackage{cite}
%\usepackage{subfigure} 
\usepackage{multirow}
\usepackage{graphicx}
\usepackage[parfill]{parskip}  
\usepackage{makeidx}
\usepackage{color}
\usepackage{lpic}
\usepackage{longtable}
\usepackage[font={scriptsize,it}]{caption}
%\newcommand{\captionfonts}{\it}
%\makeatletter
%\long\def\@makecaption#1#2{%
%\vskip\abovecaptionskip
%\sbox\@tempboxa{{\captionfonts #1: #2}}%
%\ifdim \wd\@tempboxa >\hsize
%{\captionfonts #1: #2\par}
%\else
%\hbox to \hsize{\hfill\box\@tempboxa\hfill}%
%\fi
%\vskip\belowcaptionskip}
%\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%DEFINITIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\definecolor{red}{rgb}{1,0,0} 
\def\commentr#1{\textcolor{red}{\tt [#1]}}
%\def\be{\begin{equation}}
%\def\ee{\end{equation}}

\def\H{\,{\mathcal  H}}
\def\N{\,{\mathcal N}}
\def\L{\,{\mathcal L}}
\def\fsky{\mathrm{sky}}
\newcommand{\obs}{\mathrm{obs}}
\newcommand{\pred}{\mathrm{pred}}
\newcommand{\eff}{\mathrm{eff}}
\newcommand{\intt}{\mathrm{int}}
\newcommand{\LMC}{\mathrm{LMC}}
\newcommand{\MW}{\mathrm{MW}}
\newcommand{\NGC}{\mathrm{NGC4258}}
\newcommand{\MAnd}{\mathrm{M31}}
\newcommand{\Cepheid}{\mathrm{Cepheid}}
\newcommand{\Anchors}{\mathrm{Anchors}}
\newcommand{\SNe}{\mathrm{SNe\,Ia}}
\newcommand{\km}{\mathrm{km}}
\newcommand{\magn}{\mathrm{mag}}
\newcommand{\second}{\mathrm{s}}
\newcommand{\Mpc}{\mathrm{Mpc}}
\newcommand{\MM}{{\mathcal M}}
\newcommand{\theo}{\mathrm{th}}

\newcommand{\fNL}  {f_{\rm NL}^{}}

% short-cuts
    \newcommand{\be}{\begin{eqnarray}}
    \newcommand{\ee}{\end{eqnarray}}
    \newcommand{\bea}{\begin{eqnarray}}
    \newcommand{\eea}{\end{eqnarray}}
    \newcommand{\ba}{\begin{array}}
    \newcommand{\ea}{\end{array}}

% math abbr.
    \newcommand{\ceff}{c_{\rm eff}}

% brackets:
    \renewcommand{\(}{\left(}
    \renewcommand{\)}{\right)}
    \renewcommand{\[}{\left[}
    \renewcommand{\]}{\right]}
    \newcommand{\la}{\langle}
    \newcommand{\ra}{\rangle}
    \newcommand{\lla}{\left\langle}
    \newcommand{\rra}{\right\rangle}
    \newcommand{\lcb}{\left\{}
    \newcommand{\rcb}{\right\}}

    \renewcommand{\aa}{e_\pi}
    \newcommand{\ff}{f_\pi}
    \let\mug\gg
    \renewcommand{\gg}{g_\pi}

\makeindex
\date{}                                           

%\includeonly{chapter-h0}

\includeonly{intro,chapter-ade,chapter-h0,chapter-mnu,chapter-vsk,chapter-outlook,appendix1-ade,appendix2-ade,appendix2-mnu}

\begin{document}

\renewcommand{\figurename}{\textbf{Fig.}}
\renewcommand{\tablename}{\textbf{Tab.}}

\thispagestyle{empty}
\begin{center}  

\Large  \textbf{Cosmological constraints: \\
anisotropic dark energy, the Hubble constant, and the neutrino mass} \\
  
\vspace{3.5cm}  

\large Ph. D. Thesis

\vspace{3.5cm}

submitted at the Department of Theoretical Physics \\

of the \\

UNIVERSITY OF GENEVA\\

to obtain the degree of \\

\textit{Docteur ès Sciences, mention Physique}\\

by\\

\large \textbf{Wilmar Alberto Cardona Castro}\\  

\vspace{3.5cm} 

\large Thesis N. \\  
\normalsize 2016  

\end{center}  

\newpage  
\pagenumbering{roman}
  
  
\vspace{6cm}  
\begin{center}

\large \textit{To le Sal\`{e}ve and its steepest way up}
\normalsize

\end{center}
  
\newpage  

\addcontentsline{toc}{section}{Acknowledgements}  
\chapter*{Acknowledgements} 

\vspace{3mm}

This trip started a few years ago in Palmira, Colombia. I have no words to thank the unconditional support of my family. In particular, my parents gave the right initial conditions and have been accompanying me despite the distance. When I was in Rio de Janeiro doing my master, Marcela \~{N}a\~{n}ez helped me out with bureaucracy required by my funding agency. My Colombian friends Carlos, Roc\'{i}o, Mauricio, Johanna, and Sergio encouraged me to pursue this project and made jokes at the right moment. When I was in Colombia doing my bachelor, Luis Granda and Juan Granada provided the references for my application to the scholarship I had the past four years. I thank the support of the Departamento Administrativo de Ciencia, Tecnolog\'{i}a e Innovaci\'{o}n (Colciencias) as well as the administrative help of LASPAU.  

I am pretty grateful with Martin Kunz who gave me the opportunity to come to one of the best cosmology groups in Europe. He patiently advised my work over the past four years, introduced me to research topics, encouraged me to work on different subjects, helped me with bureaucracy, and kindly provided financial aid to make my stay in Geneva easier.  

I am very thankful with Alan Heavens, Ruth Durrer, and Valeria Pettorino  for accepting to be part of the jury. I also  profited from their vast experience on the different projects where we worked jointly. Adam Riess kindly provided both Cepheid stars and Supernovae type Ia data before publication. Moreover, he answered questions about the data, suggested tests for one the projects in this thesis, and presented preliminary results in the conference COSMO-16.

Discussions with my collaborators Lukas Hollenstein, Francesco Montanari, and Marco Tucci were very helpful. I shared my office for about one year with Savvas Nesseris and profited from his experience. I thank Licia Verde and Ra\'{u}l Jimenez for hospitality during my visit to Barcelona. I appreciated the cordial reception of Bj\"{o}rn Malte and Valeria Pettorino during my visit to Heidelberg. 

My friends in Geneva played a part in this project. Indeed, Barbara Angerer was fundamental. She stayed with me and supported me in hard moments. Moreover, she carefully read several parts of the thesis. David Mogrovejo and Florian Teischinger read both introduction and summary, and made my stay in Geneva very pleasant. Rachel Gepp backed me up from Brazil. 

I am grateful with Francine Gennai-Nicole and C\'{e}cile Jaggi-Chevalley who nicely helped me with bureaucracy. Andr\'{e}as Malaspinas was always available to help with computational resources and the website administration. I am thankful to Pascale Sciarini, Stella Saavedra, Gilbert Fisher, and  Monique Vatter for helpful discussions. Finally, I thank Yves Roset who gave me the opportunity to be part of his association and read the French version of the abstract. 
  
\newpage  

\addcontentsline{toc}{section}{Abstract}  
\chapter*{Abstract}  

\vspace{3mm}

This thesis focused on cosmological constraints. First, we investigated the problem of determining the fundamental causes of the universe's accelerated expansion. We studied how the presence of a non-zero dark energy anisotropic stress impacts  both the dark matter and dark energy perturbations, as well as Cosmic Microwave Background (CMB) angular power spectrum. %which encompasses both internally and externally sourced anisotropic stress, that additionally allows for a scale dependence. %In particular, we have investigated how the presence of a non-zero dark energy anisotropic stress impacts both the dark matter and dark energy perturbations, as well as CMB angular power spectrum. We found approximate solutions for both dark matter and dark energy perturbations in some particular scenarios and constrained dark energy anisotropic stress parameters with recent data sets. 
%Gleaning information about cosmological parameters from all the available data sets will surely shed light on the shortcomings of the $\Lambda CDM$ model. Indeed, upcoming galaxy surveys such as Dark Energy Survey (DES), Dark Energy Spectroscopic Instrument (DESI), Large Synoptic Survey Telescope (LSST), Physics of the Accelerating Universe Survey (PAUS), and EUCLID will play a key role in understanding the accelerated expansion of the universe and constraint the neutrino masses. 
%The come of new astrophysical data will make necessary very careful analyses and appropriate modelling of the statistical properties of the matter density field. In particular, since upcoming galaxy surveys will probe scales comparable to the horizon, analyses must properly include relevant relativistic effects. 
Second, we investigated the impact of neglecting lensing convergence when analysing data from a EUCLID-like survey. We showed that neglecting lensing convergence when constraining neutrino masses would lead to spurious detection of their absolute mass scale, thus hindering one of the key goals of future surveys. 
%Moreover, we found that since biases of cosmological parameters in analyses neglecting lensing might reach several standard deviations, the usual linear approximation in Fisher matrix formalisms breaks down,  therefore it might no longer be appropriate. We have then adopted a Markov chain Monte Carlo (MCMC) approach to yield reliable forecasts.
%Reliable, accurate, model independent measurements of the Hubble constant $H_0$ are essential to understand the physics behind the phenomenologically successful $\Lambda CDM$ model. Accurate and precise determinations of $H_0$ will make it possible to put tighter constraints on dark energy parameters and the mass of neutrinos. 
%Although direct measurements of $H_0$ have proven to be difficult (e.g., control of systematic errors, relatively small data sets, fully consistency of different methods for measuring distances), remarkable progress has been achieved over past decades; improvements include an enlarged sample of supernovae hosts having a Cepheid calibrated distance, reduction of uncertainties on anchor distances, and increase of infra-red observations of Cepheid stars. All these efforts have yielded a direct $H_0$ measurement almost as precise as the indirect determination for the $\Lambda CDM$ model derived by the Planck collaboration. 
%The group led by Adam Riess has recently found a $H_0$ value which is in a $\approx 3\sigma$ disagreement with the one derived from CMB measurements. The reasons underlying this tension are unclear (e.g., remaining CMB systematics, issues with the applied statistical methods), but if the disagreement is proved robust, this might revolutionise physics. 
Third, we developed a statistical method using Bayesian hyper-parameters to measure the Hubble constant $H_0$ with available data. The method allows a comprehensive treatment of available data sets with no need for arbitrary outlier rejection algorithms. Our best estimate of $H_0 = 73.88 \pm 2.16 \km\, \second^{-1}\, \Mpc^{-1}$ is in good agreement with previous direct measurements, but it is slightly less precise. This can be understood since our method reveals remaining inconsistencies among the data sets. 

\newpage

\addcontentsline{toc}{section}{R\'{e}sum\'{e}}  
\chapter*{R\'{e}sum\'{e}}  

\vspace{3mm}

Le contenu de cette th\`{e}se a pour sujet les contraintes cosmologiques. Nous avons tout d'abord \'{e}tudi\'{e} les  causes fondamentales de l'acc\'{e}l\'{e}ration de l'ex\-pansion de l'univers. Nous nous sommes interrog\'{e}s sur les effets de la pr\'{e}sence d'un stress m\'{e}chanique anisotrope dans les perturbations de mati\`{e}re et d'\'{e}nergie sombres, de m\^{e}me que dans les propri\'{e}t\'{e}s statistiques du fond cosmique des micro-ondes (Cosmic Microwave Background). Dans un second temps, nous avons estim\'{e} l'importance toute particuli\`{e}re de l'effet de lentille gravitationnelle lorsque l'on analyse les donn\'{e}es d'un catalogue de galaxies tel que EUCLID. On a montr\'{e} que n\'{e}gliger cet effet dans les contraintes cosmologiques de la masse de neutrinos aurait pour cons\'{e}quence de fausser les mesures de leur masse, emp\^{e}chant ainsi une des r\'{e}sultats essentiels des futurs catalogues de galaxies. Dans une troisi\`{e}me, nous avons d\'{e}velopp\'{e} une m\'{e}thode statistique qui emploie des hyperparam\`{e}tres Bayesiens (Bayesian hyper-parameters) afin de mesurer le param\`{e}tre Hubble $H_0$ avec les donn\'{e}es disponibles. Cette approche permet un traitement compr\'{e}hensif des ensembles de donn\'{e}es disponibles sans l'utilisation d'algorithmes arbitraires pour le traitement de donn\'{e}es aberrantes. Notre meilleur mesure de $H_0 = 73.88 \pm 2.16 \km\, \second^{-1}\, \Mpc^{-1}$ correspond parfaitement avec les mesures pr\'{e}c\'{e}dentes. N\'{e}anmoins, elle a une incertitude plus grande. Ceci s'explique par le fait que la m\'{e}thode utilis\'{e}e met en \'{e}vidence des incoh\'{e}rences entre les ensembles de donn\'{e}es. 

\tableofcontents

%\listoffigures

%\listoftables 

\newpage

\pagenumbering{arabic}

\include{intro}

%\include{chapter1}

\include{chapter-ade}

\include{chapter-mnu}

\include{chapter-h0}

\include{chapter-vsk}

\include{chapter-outlook}

\addcontentsline{toc}{chapter}{Appendices}  
\chapter*{Appendices}  

\appendix

\include{appendix1-ade}

\include{appendix2-ade}

%\include{appendix1-h0}

%\include{appendix1-mnu}

\include{appendix2-mnu}
%
\bibliography{H0_refs}
\bibliographystyle{JHEP}
%


\end{document}  